---
title: "Grocery Analysis"
output:
  word_document: default
  html_document: default
date: "2024-11-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


https://github.com/fivethirtyeight/data/tree/master/police-killings


```{r}
library(readr)
police_data <- read_csv("police_killings.csv")

```

```{r}
# Check the structure of the dataset
str(police_data)

# Check for missing values
summary(police_data)

# Check the column names
colnames(police_data)

```

## Cleaning

```{r}
# Convert columns to numeric where necessary
police_data$age <- as.numeric(police_data$age)
police_data$share_white <- as.numeric(police_data$share_white)
police_data$share_black <- as.numeric(police_data$share_black)
police_data$share_hispanic <- as.numeric(police_data$share_hispanic)
police_data$p_income <- as.numeric(police_data$p_income)
police_data$h_income <- as.numeric(police_data$h_income)

# Check for missing values again after conversion
summary(police_data)

# Handle missing values - For example, remove rows with missing income data
police_data_clean <- police_data[!is.na(police_data$h_income), ]



```

```{r}
# Basic EDA: Check the distribution of raceethnicity
table(police_data_clean$raceethnicity)


```

```{r}
# Investigate cause of death
table(police_data_clean$cause)


```

```{r}
# Explore the income distribution
hist(police_data_clean$h_income, main = "Income Distribution", xlab = "Household Income", col = "blue", breaks = 20)
```


# Research Questions



### 1. **Geographic Patterns and Police Killings** (Exploratory Analysis)
   - **RQ4**: *Is there a regional variation in the number of police killings across different U.S. states?*
     - **Analysis Approach**: 
       - **Descriptive Statistics**: Calculate the total number of police killings by state and plot the distribution.
       - **Visualizations**: Create bar charts or maps to visualize the frequency of police killings across states.
       - **Chi-Square Test**: To check if the distribution of police killings is independent of regions or states.
       - **Key Variables**: `state`, `year`, `name` (for counts), `gender`, `raceethnicity`.

### 2. **Socio-Economic Factors and Police Killings** (Regression Analysis)
   - **RQ9**: *Is there a relationship between income levels (e.g., personal, household, or county income) and the rate of police killings in different areas?*
     - **Analysis Approach**:
       - **Regression Analysis**: Perform a linear regression to analyze the relationship between income and the rate of police killings in a region.
       - **Variables**:
         - **Independent Variables**: `p_income` (personal income), `h_income` (household income), `county_income` (county income).
         - **Dependent Variable**: `pop` (population of the area or the number of police killings).
         - **Control Variables**: `raceethnicity`, `state`, `armed`.
       - **Model**: Use a multiple linear regression model to estimate how income levels (individual, household, or county) influence the frequency of police killings, controlling for other factors like race and whether the individual was armed.
       - **Hypothesis**: Higher income areas will have fewer police killings due to different levels of law enforcement resources, socioeconomic stability, or community-police relationships.



## Summary statistics

```{r}
# Example: Assuming 'data' is a dataframe with a column 'state'
# If you haven't defined 'data' yet, create a dummy dataframe for testing


# List of left-wing (Democratic-leaning) and right-wing (Republican-leaning) states
left_wing_states <- c("CA", "NY", "IL", "WA", "OR", "CO", "MA", "MI", "PA", "MN", "VA", "NJ", "MD", "CT", "RI", "VT", "HI")
right_wing_states <- c("TX", "FL", "GA", "OH", "AZ", "NC", "SC", "WI", "MO", "IN", "KY", "LA", "AL", "OK", "TN", "MS", "AR")

# Create a new variable 'state_category' based on the political leaning
police_data_clean$state_category <- ifelse(police_data_clean$state %in% left_wing_states, "Left-Wing", 
                              ifelse(police_data_clean$state  %in% right_wing_states, "Right-Wing", "Unaffiliated"))


```



```{r}
library(gtsummary)

library(gtsummary)

# Example: Summarizing police killings dataset with key variables (age, gender, raceethnicity, income)
table2 <- police_data_clean |> 
  tbl_summary(
    include = c(age,gender,  raceethnicity, p_income, h_income, county_income),  # Variables to summarize
    by = state_category,  # Split the table by state
    missing = "no"  # Do not list missing data separately
  ) |> 
  add_n() |>  # Add column with total number of non-missing observations
  add_p() |>  # Test for a difference between groups (state in this case)
  modify_header(label = "**Variable**") |>  # Update column header
  bold_labels()  # Bold labels for clarity

# View the summary table
table2


```









```{r}
# Fit a full model with all relevant predictors
full_model <- lm(pop ~ p_income + h_income + county_income + raceethnicity + armed + state_category, data = police_data_clean)

# Perform stepwise regression (both forward and backward)
stepwise_model <- step(full_model, direction = "both")

# Summary of the final model after stepwise selection
summary(stepwise_model)

```


```{r}
# Residuals vs Fitted values plot
plot(stepwise_model, which = 1)

# Normal Q-Q plot for residuals
plot(stepwise_model, which = 2)

# Check for multicollinearity using VIF (Variance Inflation Factor)
library(car)
vif(stepwise_model)

# Check residuals for normality
shapiro.test(resid(stepwise_model))

```

